---
title: "Final Review"
subtitle: "EC 320: Introduction to Econometrics"
author: "Tami Ren"
date: "Fall 2022"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    # self_contained: true
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, middle

```{r Setup, include = F}
options(htmltools.dir.version = FALSE)
library(pacman)
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, here, tidyverse,
  emoGG, latex2exp, ggplot2, ggthemes, viridis, extrafont, gridExtra,
  kableExtra,
  data.table, dagitty, ggdag,
  dplyr, gganimate,
  lubridate,
  magrittr, knitr, parallel, xaringanthemer
)
# Define colors
red_pink <- "#e64173"
met_slate <- "#23373b" # metropolis font color
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  #dpi = 300,
  #cache = T,
  warning = F,
  message = F
)  
theme_simple <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 14),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  legend.position = "none"
)
theme_gif <- theme_bw() + theme(
  axis.line = element_line(color = met_slate),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  text = element_text(family = "Fira Sans", color = met_slate, size = 14),
  axis.text.x = element_text(size = 12),
  axis.text.y = element_text(size = 12),
  axis.ticks = element_blank()
)

style_duo_accent(
  primary_color = "#1F4257",
  secondary_color = "#F97B64",
  inverse_header_color = "#FFFFFF"
)

```

# Prologue

---
# Housekeeping

## Final Exam

Tomorrow! On Canvas. At 12:00 Pm. 

- I will be on Zoom to answer any questions

## Problem Set 4

Due Friday at Midnight 

## Office Hours

- No office hours Friday morning
- Holding extra office hours Thursday evening: Available via Zoom from 6-8 pm PST 


---
class: inverse, middle

# Final Exam

---
# Final Exam Topics

.green[Anything from the lectures, labs, or problem sets --- the focus will be on the latter half of the course. The former half of the course is the foundational for the latter half, so still know it] 

---
# Final Exam Topics

Calculating treatment effect 

Calculating coefficient

Goodness of fit

Hypothesis testing

Confounding Variables

.pink[Omitted-variable bias]

- Know when omitting a variable causes bias
- Provide examples of problematic omitted variables

Controls: Bad and Multicollinearity 
---
# Final Exam Topics

Making predictions using fitted regression model

- *e.g.,* using a Hedonic model


Multiple Linear Regression: Inference

- *F* tests (multiple parameters)

 - State null hypothesis
 - Identify restricted and unrestricted models
 - Calculate *F* statistic
 - Use table to find *F*.sub[*crit*]
 - *F* .mono[>] *F*.sub[*crit*]?
 - State conclusion of the test

*t* tests (single parameter)


---
# Final Exam Topics

Confidence intervals

- Formula, interpretation, and comparison of different intervals for the same coefficient


---
# Final Exam Topics

Categorical Variables: How do you interpret coefficients on binary variables?

- **Note:** Depends on the presence of interaction terms and whether the outcome variable is transformed

Dummy variable trap

What is the reference category?

- How do you back out group-specific averages from a dummy variable regression?
- How do coefficient estimates change when you change the reference category?

---
# Final Exam Topics

Interactive Relationships: How do you interpret interaction coefficients?

- Binary $\times$ binary
- Binary $\times$ continuous
- Continuous $\times$ continuous

How does an interaction term change how you interpret the effect of the variable of interest on the outcome variable?

- Marginal effects (partial derivative)

---
# Final Exam Topics

Nonlinear Relationships: Identify nonlinear models

- OLS can handle nonlinear variables, but not nonlinear parameters

Transform nonlinear models


---
# Final Exam Topics

## Nonlinear Relationships: How do you interpret coefficients in the presence of logarithmic transformations?

- Level $Y$, level $X$
- Level $Y$, log $X$
- Log $Y$, level $X$
- Log $Y$, log $X$


**Quadratic models:** interacting $X$ with itself

- Calculate marginal effects to understand how $X$ affects $Y$

---
# Final Exam Structure

Same format as Midterm --- may have 1 or 2 more questions 


---
class: inverse, middle

# Practice

---
# .mono[gapminder] Package

Data on population, GDP per capita, and life expectancy 

Unit of observation: country-year 

- All countries, every 5.super[th] year between 1957 and 2007

```{r}
p_load(gapminder)
data <- get('gapminder')
head(data)
```

---
# GDP per Capita by Continent

```{r}
reg <- lm(gdpPercap ~ continent, data = data)
tidy(reg)
nobs(reg)
summary(reg)$r.squared
```

---
# GDP per Capita by Continent

```{r}
reg2 <- lm(log(gdpPercap) ~ continent, data = data)
tidy(reg2)
nobs(reg2)
summary(reg2)$r.squared
```

---
# GDP per Capita by Continent

```{r}
data <- data %>% 
  mutate(continent = ifelse(as.character(continent) %in% 
                              c("Europe", "Americas", "Oceania"), 
                            "The West",
                            as.character(continent)))
reg3 <- lm(log(gdpPercap) ~ continent, data = data)
tidy(reg3)
summary(reg3)$r.squared
```

---
# GDP per Capita by Continent

**Null hypothesis:** Europe, the Americas, and Oceania have the same average GDP per capita.

--

**Setup *F* Test**

```{r}
# restricted model: reg3
# unrestricted model: reg2

# get r-squared from restricted and unrestricted models
r2_restrict <- summary(reg3)$r.squared
r2_unrestrict <- summary(reg2)$r.squared

# number of resrictions
n_restrict <- 2 

# degrees of freedom in unrestricted model
dof <- summary(reg3)$df[2] 

# significance level
alpha <- 0.01
```

---
# GDP per Capita by Continent

**Null hypothesis:** Europe, the Americas, and Oceania have the same average GDP per capita.

**Conduct *F* Test**

```{r}
# calculate F stat:
f_stat <- ((r2_unrestrict - r2_restrict)/n_restrict)/((1 - r2_unrestrict)/dof) 

# find critical value of F:
f_crit <- qf(1-alpha, n_restrict, dof)

# reject null if:
f_stat > f_crit
```

--

**Conclusion:** At least one of the continents in "The West" has a different average GDP per capita than the others.

---
class: white-slide
.center[**Association of GDP per Capita with Life Expectancy**]
```{R, echo = F, escape = F}
tab <- data.frame(
  v1 = c("log(GDP/Capita)", "", "log(GDP/Capita) ? Americas", "", "log(GDP/Capita) ? Asia", "", "log(GDP/Capita) ? Europe", "", "log(GDP/Capita) ? Oceania", "", "Continent Dummies?", "Observations", "R-Squared"),
  v2 = c(6.42, "(0.183)", "", "", "", "", "", "", "", "", "Yes", 1704, 0.704),
  v3 = c(5.69, "(0.325)", 4.03, "(0.674)", 0.561, "(0.421)", 0.614, "(0.597)", 5.20, "(4.36)", "Yes", 1704, 0.710),
  v4 = c(0.112, "(0.004)", "", "", "", "", "", "", "", "", "Yes", 1704, 0.665),
  v5 = c(0.111, "(0.006)", 0.047, "(0.013)", -0.002, "(0.008)", -0.019, "(0.012)", 0.035, "(0.084)", "Yes", 1704, 0.669)
) %>% kable(
  escape = F,
  col.names = c("", "Life Expectancy", "Life Expectancy", "log(Life Expectancy)", "log(Life Expectancy)"),
  align = c("l", rep("c", 5))
) %>%
row_spec(1:13, color = met_slate) %>%
row_spec(seq(2,10,2), color = "#c2bebe") %>%
row_spec(1:13, extra_css = "line-height: 110%;") %>% 
column_spec(1, color = "black", italic = T) %>% 
kable_styling(font_size = 18.2)
tab %>% column_spec(2, bold = T)
```

---
class: white-slide
.center[**Association of GDP per Capita with Life Expectancy**]
```{R, echo = F, escape = F}
tab %>% column_spec(3, bold = T)
```

---
class: white-slide
.center[**Association of GDP per Capita with Life Expectancy**]
```{R, echo = F, escape = F}
tab %>% column_spec(4, bold = T)
```

---
class: white-slide
.center[**Association of GDP per Capita with Life Expectancy**]
```{R, echo = F, escape = F}
tab %>% column_spec(5, bold = T)
```