---
title: "Week 1 Worksheet"
author: "Tami Ren"
date: "6/19/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. A random variable X is defined to be the difference between the higher values and the lower value when two dice are thrown. If they have the same value, X is defined to be zero. 

a) Find the probability distribution for X 

b) Find the expected value of X 

c) Calculate $E(X^2)$

d) Calculate the population variance and the standard deviation of X

e) Find the variance of the random variable X in (4) using the equation $\sigma_x^2 = E(X^2) - \mu_x^2$. Does it equal your answer from part (d)? 






2. Prove that if Y=b, where b is a constant, COV(X,Y) = 0









3. Prove if Y=V+W, var(Y) = var(V) + var(W) + 2cov(V,W)









4. Consider data gathered on adult domestic cats' body weight and heart weight: 

```{r, echo=FALSE}

cat = c(1:5)
body_weight = c(2,2.2,2.4,2.4,3.6)
heart_weight = c(6.5,7.6,9.1,7.3,11.8)

cbind(cat,body_weight,heart_weight)
```
a) What is the sample mean of Body Weight 
b) What is the sample variance of Heart Weight
c) What is the sample correlation coefficient between Body Weight and Heart Weight 

