---
title: "Lab 3"
author: "Tami Ren"
date: "July 2 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE)

options(htmltools.dir.version = FALSE)
library(pacman)
p_load(broom, latex2exp, leaflet, ggplot2, ggthemes, viridis, dplyr, magrittr, knitr, parallel, rddtools, readxl, dslabs, gapminder, extrafont, Ecdat, wooldridge, tidyverse, janitor, kableExtra, gridExtra, estimatr, jtools, huxtable)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#3b3b9a"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
slate <- "#314f4f"
# Dark slate grey: #314f4f
# Notes directory
dir_slides <- "/Users/tami/Dropbox (University of Oregon)/EC320--Ren--2022/Homework/Homework2"
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 4,
  fig.width = 6,
  # dpi = 300,
  # cache = T,
  warning = F,
  message = F
)
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  axis.text.x = element_text(size = 10),
  axis.text.y = element_text(size = 10),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title.x = element_text(angle = 0, vjust = 0.5),
  axis.title.y = element_text(angle = 90, vjust = 0.5),
  legend.position = "none",
  axis.line = element_line(color="black", size = .5)
)
```

# {.tabset .tabset-fade .tabset-pills}

## Lab 3 

In this lab, I will show you how to run a regression, create a pretty regression table, and evaluate results from a simulation. 

### The data-generating process

Suppose the effect of variable X on Y in the population is defined by the following data generating process (DGP). The distribution of X is normal with a mean of 10 and a variance of 2. 

$$
\begin{align}
  \text{X}\sim&\mathop{\text{Norm}}(10, 2), \\
  \text{Y} =& \ \ 1 + 5\text{X} + \varepsilon, \ \\
\end{align}
$$

where $\varepsilon\sim\mathop{N}(0,1)$ 

```{r, ssim_dgp}
library(pacman)
p_load(dplyr, ggplot2)
# Choose an observation count (number of rows)
n <- 100000
set.seed(12345)
# Generate data in a tibble
df = tibble( 
  e = rnorm(n, sd = 1),
  X = rnorm(n, mean = 10, sd=2),
  Y = 1 + 5*X + e
)
```


Suppose you are estimating the effect of X on Y. As an economist, you need  a sample of data. You are able to get a sample of data for 100 different people in the population. 

```{r sample}
#set.seed(145)
sample_df =df[sample(nrow(df), size=100),]
```

### Regression



```{r,regression}
reg1 = lm(Y ~ X, sample_df)

```
#### 2. Present your result in a table (use stargazer or export_summs) 

```{r, one_table}
library(stargazer)
#stargazer(reg1 ,keep=c("Y","X"), type="html", covariate.labels="X",dep.var.labels="Y",digits=3)
```
<table style="text-align:center"><tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left"></td><td><em>Dependent variable:</em></td></tr>
<tr><td></td><td colspan="1" style="border-bottom: 1px solid black"></td></tr>
<tr><td style="text-align:left"></td><td>Y</td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">X</td><td>4.998<sup>***</sup></td></tr>
<tr><td style="text-align:left"></td><td>(0.054)</td></tr>
<tr><td style="text-align:left"></td><td></td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">Observations</td><td>100</td></tr>
<tr><td style="text-align:left">R<sup>2</sup></td><td>0.989</td></tr>
<tr><td style="text-align:left">Adjusted R<sup>2</sup></td><td>0.989</td></tr>
<tr><td style="text-align:left">Residual Std. Error</td><td>1.048 (df = 98)</td></tr>
<tr><td style="text-align:left">F Statistic</td><td>8,699.575<sup>***</sup> (df = 1; 98)</td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left"><em>Note:</em></td><td style="text-align:right"><sup>*</sup>p<0.1; <sup>**</sup>p<0.05; <sup>***</sup>p<0.01</td></tr>
</table>


#### Residuals

```{r, residuals}

plot(residuals(reg1))

sample_df =sample_df %>% mutate(diff_err = residuals(reg1) - sample_df$e,
                                resid = residuals(reg1))

#View(sample_df)

sample_df %>% ggplot() + 
  geom_density(aes(diff_err))

```
#### Simulations

```{r, sim-fit_do-not-change}

sim_fit_fun <- function () {
  fit <-lm(Y ~ X, data=df, subset = sample(nrow(df), 1000))
  coef(fit)
  }

sim_fit_df <- as.data.frame(t(replicate(10000, sim_fit_fun())))

#View(sim_fit_df)

```

### Plot the simulation

```{r, density_sim}

sim_fit_df %>% ggplot() + 
  geom_density(aes(X), fill="dark blue") + theme_minimal() + labs(title = "Density of the coefficient values from the simulation ")

mean(sim_fit_df$X)

```
